# Introdiction to Statistics {#sec-intro}

```{r}
#| label: load_packages
#| include: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(tsibble)
library(feasts)
library(fable)
```

## What is Statistics?

The science of collecting, organizing, and analyzing data, and presenting and interpreting the results derived from that data to gain insights and make informed decisions.

## What is Biostatistics?

When statistical methods are applied to specific fields, the discipline often takes on a specialized name. When statistical methods are applied to biological and medical sciences, it is called Biostatistics. Table 1 shows Specialized fields where statistics is applied.


---
title: "Specialized Fields of Statistics"
format: html
---

```{r, tab.cap="Specialized fields where statistics is applied"}
library(knitr)
library(kableExtra)

# Fields and descriptions
fields <- c(
  "Epidemiology",
  "Astrostatistics",
  "Biostatistics",
  "Demography",
  "Social statistics",
  "Chemometrics",
  "Econometrics",
  "Psychometrics",
  "Pharmacometrics",
  "Bioinformatics / Biostatistics in Molecular Biology",
  "Agrimetrics / Agricultural Statistics",
  "Ecological Statistics / Biostatistics in Ecology",
  "Actuarial Statistics",
  "Forensic Statistics",
  "Spatial Statistics",
  "Jurimetrics",
  "Sports Analytics / Sports Statistics",
  "Environmental Statistics"
)

descriptions <- c(
  "The study and analysis of the patterns, causes, and effects of health and disease conditions in populations.",
  "Applies statistical analysis to the understanding of astronomical data.",
  "Studies biological phenomena and provides statistical methods for analysis of biological data.",
  "Statistical study of all populations, including size, distribution, and structure.",
  "Study of human behavior in a social environment using statistical methods.",
  "Science of extracting information from chemical systems by data-driven means.",
  "Uses economic theory, mathematics, and statistical inference to quantify economic phenomena.",
  "Applies statistical methods to psychological measurements, tests, and surveys.",
  "Modeling drug dose-response relationships, pharmacokinetics, and pharmacodynamics.",
  "Analyzes DNA, RNA, protein data, sequence alignments, and gene expression patterns.",
  "Statistics applied to agriculture — crop yield trials, soil studies, plant breeding, pest control.",
  "Statistics applied to ecology and wildlife studies — population surveys, species distribution, biodiversity indices.",
  "Discipline that deals with assessing the risks in insurance and finance.",
  "Applies statistics to analyze forensic data, including DNA testing results.",
  "Analysis of spatially referenced data to detect patterns or relationships.",
  "Application of probability and statistics to law and legal decisions.",
  "Statistics applied to sports — analyzing player performance, game strategies, and injury risks.",
  "Statistics in environmental science — air/water quality, climate data, pollution modeling, conservation research."
)

# Create data frame
df <- data.frame(Field = fields, Description = descriptions)

# Render table using kableExtra
kable(df, format = "html", booktabs = TRUE, caption = "Specialized fields where statistics is applied") %>%
  kable_styling(full_width = TRUE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```

## Role of statistics in zoology and biological sciences

Statistics play a vital role in every aspect of zoological research, helping scientists design studies, analyze data, and make informed conclusions.

Examples in zoology include:

- Population estimation – Using sampling and statistical models to estimate the number of elephants in a forest or birds in a wetland.

- Morphometric analysis – Comparing body size, wing length, or foot circumference among different species or populations.

- Behavioral studies – Analyzing patterns such as feeding,  or migration behaviors using observational data.

- Ecological modeling – Studying the relationships between species distribution and environmental factors like habitat type, climate, or human disturbance.

- Conservation planning – Assessing risk factors for endangered species and predicting population trends using statistical models.


- Experimental studies - effects of temperature on insect development.


