# Descriptive Statistics {#sec-chap5}

## Dataset Use for Illustrations

We will use `penguins` dataset in the `palmerpenguins` package.

Variable description

```{r}
colnames <- colnames(penguins)
description <- c("penguin species (Adélie, Chinstrap and Gentoo)", 
                 "island in Palmer Archipelago, Antarctica (Biscoe, Dream or Torgersen)", 
                 "bill length (millimeters)", 
                 "bill depth (millimeters)", 
              "flipper length (millimeters)",
              "body mass (grams)",
              "penguin gender (female, male)",
              "the study year (2007, 2008, or 2009)")
df <- data.frame(colnames, description)
colnames(df) <- c("Variable name", "Variable description")
df |> knitr::kable()
```

Preview of the dataset

```{r}
library(palmerpenguins)
penguins
```


## Qualitative data: Univariate Analysis

Chart types we can use: Pie chart, Bar chart

Tables: Tables with counts and percentages




## Qualitative data: Bivariate Analysis

## Quntitative data: Univariate Analysis

## Quantitative data: Bivariate Analysis

## Qualitative vs Quantitative

## Quantitative vs Quantitative

## Quantitative vs Quantitative with a Qualitative variable